---
title: "`data.table` Case Study - Women in Parliament"
author: "Saghir Bashir"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load all the libraries that you will use once here
library(here)
library(data.table)
library(countrycode)
library(ggplot2)
library(magrittr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Changing default settings here.
theme_set(theme_classic(base_size = 16))
```

# Case Study: Women in Parliament

# Objectives

*Explore the geographical and time trends for the percentage of women in 
national parliaments.*

## Questions

The objective is to look at the geographical and time trends in the data. We will
answer the following questions.

+ What are the time trends for Portugal?
+ How does Portugal compare to other countries?
+ Which countries have the highest percentage of women in parliament by year?
+ How do continents compare?
+ What are the global trends over time?

# Data Details

The raw data for *"Proportion of seats held by women in national parliaments"* 
includes the percentage of women in parliament (_"single or lower parliamentary chambers 
only"_) by country (region) and year. It can be downloaded from:

- https://data.worldbank.org/indicator/SG.GEN.PARL.ZS 

As part of its "open data" mission the World Bank offers _"free and open access 
to global development data"_ kindly licensed under the "Creative Commons Attribution 
4.0 (CC-BY 4.0)".

## Source Data

The data originates from the "Inter-Parliamentary Union" (IPU) which
provides an *_"Archive of statistical data on the percentage of women in 
national parliaments"_* going back to 1997 on a monthly basis:

- http://archive.ipu.org/wmn-e/classif-arc.htm

The World Bank data is for “single or lower parliamentary chambers only”, while 
the IPU also presents data for “Upper Houses or Senates”. Moreover, the IPU provides 
the actual numbers used to calculate the percentages (which the World Bank does not).

## Data limitations

Take caution when interpreting these data, as parliamentary systems vary from country 
to country, and in some cases over time. Some of the issues to consider include:

- Who has, and who does not have, the right to become a Member of Parliament (MP)?
- How does someone become an MP? Through democratic elections? How is "democratic
election" defined? 
- What is the real power of MPs and their parliament? Can MPs make a difference?

## Data definitions & assumptions

### "Women"

The definition for “women” is not given, so we will assume that it refers to a 
binary classification for gender (sex).

### "Country (Region)"

The definition of countries and regions can change over time.  (e.g. formation of new
countries after conflicts, new member states joining a pre-existing collective). 
How are these changes reflected in the data? How do they affect the interpretation? 

For example, the “European Union” has changed over time. It started in the 1950s as 
a block of six European countries (known as the “European Community”) and has expanded 
over the years to 28 countries (with the United Kingdom about to depart). This raises 
the question of how the European Union average is calculated. For a given year, is it 
calculated based on the actual member states in that year or on all of the current 
member states?

```{r dataSource, message=FALSE, echo=FALSE}
wbCSVs <- list.files(here("data"), full.names = FALSE, pattern = "[.]csv$")
metaInd <- wbCSVs[grepl("indicator", wbCSVs, ignore.case = TRUE)]
indInfo <- fread(here("data", metaInd), 
                   header = TRUE)
indName <- indInfo[, INDICATOR_NAME]
```

## World Bank Datasets

The World Banks provides the following data files.
```{r datasetsWB, echo=TRUE}
wbFiles <- list.files(here("data"), full.names = FALSE)
indData <- wbFiles[grepl("^API", wbFiles)]
indMeta <- wbFiles[grepl("Indicator", wbFiles)]
ctyMeta <- wbFiles[grepl("Country", wbFiles)]
```

+ "**`r indData`**" contains the "`r indName`" data.
+ "**`r indMeta`**" contains the indicator meta data which is as described above. 
+ "**`r ctyMeta`**" contains the Country related meta data that can be useful to
understand the limitations of the data by country.


# Data Processing

The World Bank's "Women in Parliament" data are imported and undergo some "tidying"
before being used for the analyses in the sections below.

```{r readData, warning=FALSE}
wip <- fread(here("data", indData), 
            skip = 4, header = TRUE,
            check.names = TRUE)

# Add Continent to WiP dataset.
cl <- as.data.table(codelist)[, .(continent, wb)]

wipcl <- merge(wip, cl, by.x = "Country.Code", by.y = "wb", all.x = TRUE)
setnames(wipcl, c("continent"), c("Continent"))

WP <- melt(wipcl,
           id.vars = c("Continent", "Country.Name", "Country.Code", "Indicator.Code"),
           measure = patterns("^X"),
           variable.name = "YearC",
           value.name = c("pctWiP"),
           na.rm = TRUE)

# Note cLabel will be used to label the ends of lines in plots.
WP[, `:=`(Year = as.numeric(gsub("^X", "", YearC)), Ratio = (100 - pctWiP)/pctWiP)][
  , fYear := factor(Year)][
  , maxYear:=max(Year), by = .(Country.Name)][
    Year==maxYear, cLabel:=Country.Name][
      , c("maxYear", "YearC"):=NULL]

setcolorder(WP, c("Indicator.Code", "Continent", "Country.Name", "Country.Code", 
                  "Year", "pctWiP", "Ratio"))
setkeyv(WP, c("Indicator.Code", "Continent", "Country.Name", "Country.Code", "Year"))

WP
```

# Analysis


## Portugal - Trends and Comparisons

> What are the time trends for Portugal?

Let's start by looking at a plot of how Portugal is performing over time.

```{r PTplot, warning=FALSE}
WP[Country.Name %in% "Portugal"] %>% 
ggplot(aes(Year, pctWiP)) +
  geom_line(colour = "blue") + 
  geom_point(colour = "blue") +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0, 50)) +
  ggtitle("Portugal") +
  xlab("Year") +
  ylab("% Women in Parliament") 
```

**Interpretation**

In 1990 Portugal had 7.6% women in parliament (i.e. 12.2 men for each woman), which 
increased to 34.8% (i.e. 1.87 men for each woman) in 2018. This still falls short 
of 50% (i.e. point of gender parity in parliament).

Let's now plot how Portugal compares with other European countries. As a reference
the European Union and World averages. Finland, Hungary, Sweden and Romania are highlighted
for discussion.

```{r PTvsEurope, warning=FALSE}
WP[Continent == "Europe"] %>%
  ggplot(aes(Year, pctWiP, group=Country.Name, colour = Country.Name, label = cLabel)) +
  geom_line(colour="grey90") +
  geom_line(data = WP[Country.Name %in% c("Portugal", "European Union", "World", 
                                          "Sweden", "Finland", "Hungary", "Romania")],
            aes(colour = Country.Name)) +
  scale_x_continuous(breaks=seq(1990, 2024, 5)) +
  scale_y_continuous(limits=c(0, 50), breaks=seq(0, 50, by=10)) +
  expand_limits(x = 2024) +
  geom_label_repel(data = WP[Country.Name %in% c("Portugal", "European Union", "World", 
                                                 "Sweden", "Finland", "Hungary", "Romania")],
                   xlim = c(2018, 2024)) +
  theme(legend.position = "none") +
  ggtitle("Portugal compared to European countries") +
  xlab("Year") +
  ylab("% Women in Parliament")

```

**Interpretation**

Portugal has had more women in parliament than the European Union average since 2007
and since around 2000 compared to the world average. Hungary and Romania both had 
a higher percentage of women in parliament in 1990 (around the end of the Cold War) 
than they have had since. The key point to note is that none of these countries 
reaches equality between males and females in parliament, although Sweden and 
Finland come closest.


## Highest Percentages 

> Which countries have the highest percentage of women in parliament by year?  How do continents compare?

```{r highestContinent, collapse=TRUE}
hgWiP <- WP[!is.na(Continent)][
  order(Continent, Year, -pctWiP), head(.SD, 1), by = .(Continent, Year)][
  , CountryWiP := sprintf("%3.1f%% - %s", pctWiP, Country.Name)][
  , .(Continent, Year, CountryWiP)]

hgCont <- dcast(hgWiP, Year ~ Continent, value.var = "CountryWiP") 

hgCont %>% 
  kable(align="clllll") %>% 
  kable_styling(bootstrap_options = "striped")
```

## No Women in Parliament

Which countries have no (0%) women in parliament?

```{r zeroWiP, collapse=TRUE}
zeroWiP <- WP[!is.na(Continent) & pctWiP==0]

zeroYear <- zeroWiP[order(Country.Name), .(Year, Country.Name)][
  , toString(paste0('"', Country.Name, '"')), by = .(Year)][order(Year)]
setnames(zeroYear, c("V1"), c("zeroWiP"))

zeroYear %>% 
  kable(align="ll", col.names = c("Year", "Countries without any Women in Parliament")) %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  column_spec(1, bold = T)
```


## Global Trends

> What are the global trends over time?

In the following each country has been plotted as a line and the world "average" is
highlighted in blue. 

```{r globalTrendsLines, warning=FALSE}
WP[!is.na(Continent)] %>%
  ggplot(aes(Year, pctWiP, group = Country.Name, label = cLabel)) +
  geom_line(colour="grey90") +
  geom_line(data=WP[Country.Name=="World"], colour="blue") +
  expand_limits(x = 2021) +
  geom_label_repel(data = WP[Country.Name=="World"], 
                   xlim = c(2018, 2021), colour = "Blue") +
  theme(legend.position = "none") +
  scale_x_continuous(breaks=seq(1990, 2021, 5)) +
  scale_y_continuous(limits=c(0, 70), breaks=seq(0, 70, by=10)) +
  ggtitle("Women in Parliament: Global Trends") +
  ylab("% Women in Parliament")
```

A box whisker plot is better suited to present the variation (uncertainty) over
time.

```{r globalTrendsBoxplot, message=FALSE}
WP[!is.na(Continent)] %>%
  ggplot(aes(fYear, pctWiP)) +
  geom_boxplot(width = 0.25, outlier.size = .5) +
  scale_x_discrete(breaks=seq(1990, 2020, 2)) + 
  scale_y_continuous(limits=c(0, 70), breaks=seq(0, 70, by=10)) +
  ggtitle("Box Whisker Plot for countries") +
  ylab("% Women in Parliament") +
  xlab("Year")
```

**Interpretation**

Although the world average is going up there is still large variation between countries. 
Note that in earlier years fewer countries provided data so there is likely to be some
bais in the plots above.

# Conclusions

```{r worldConclusions, echo=FALSE}
pctWiPLast <- WP[Country.Name=="World"][order(Year), .SD[.N]][, pctWiP]
ratioWiPLast <- WP[Country.Name=="World"][order(Year), .SD[.N]][, Ratio]
yearLast <- WP[Country.Name=="World"][order(Year), .SD[.N]][, Year]
```

In `r yearLast`, globally there were `r sprintf("%3.1f%%", pctWiPLast)` women in parliament, 
i.e, for every woman in parliament there was `r sprintf("%2.1f", ratioWiPLast)` men 
in parliament. Whichever way we look at the data the conclusions is that, in general, 
the percentage of women in parliament is increasing but that gender parity in most 
parliaments is still far-off.

<hr>

# Session Information

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

